---
title: "Diamond Plots"
author: "Gjalt-Jorn Ygram Peters"
date: "`r format(Sys.time(), '%Y-%m-%d at %X');`"
output: html_document
---

```{r setup, include=FALSE}

########################################################################
### Paths
########################################################################

### Add any relevant paths to this vector. The script will select the
### correct path itself.

basePathVector <- c('B:/Data/research',
                    '');

########################################################################
### Set the variables with the paths
########################################################################

### Check which paths exist and set the first existing path in the list
### as the base path
basePath <- basePathVector[sapply(basePathVector, dir.exists)][1];

### Set the additional paths
outputPath <- file.path(basePath, 'determinant-importance');
dataPath <- file.path(basePath, 'party panel', '15.1', 'data');
dataFileName <- 'survey_652829_R_data_file.csv';
dataLoadScriptName <- 'survey_652829_R_syntax_file.R';

###########################################################
### Installing & loading the required packages
###########################################################

### If not present yet, install the
### userfriendlyscience package
if (!is.element('userfriendlyscience', installed.packages()[,1])) {
  ### Use Belgian https mirror; replace with NULL to choose
  chooseCRANmirror(ind=7);
  install.packages('userfriendlyscience', dependencies=TRUE);
}

### Load userfriendlyscience
require('userfriendlyscience');

### Load the rest of the packages, installing them if need be
safeRequire('ggplot2');
safeRequire('grid');
safeRequire('data.tree');
safeRequire('pander');

###########################################################
### R Markdown configuration
###########################################################

### Setting default knitting options
knitr::opts_chunk$set(echo=FALSE);
knitr::opts_chunk$set(comment=NA);
knitr::opts_chunk$set(dev="png", 
		  		            dev.args=list(type="cairo"),
			    	          dpi=100);
knitr::opts_chunk$set(fig.width=5);
knitr::opts_chunk$set(fig.height=5);
knitr::opts_chunk$set(cache=FALSE);

options(scipen=100);

panderOptions('table.split.table', Inf);
panderOptions('table.alignment.rownames', "left");

# options(figure_counter = TRUE);
# options(table_counter = TRUE);

setFigCapNumbering();
setTabCapNumbering();

```

```{r preparation}

########################################################################
### Load data, extract, translate, and apply variable names and labels,
### set determinant structure, and compute scales
########################################################################

### Import data from LimeSurvey data file
dat <- importLimeSurveyData(datafile = file.path(dataPath, dataFileName),
                            scriptfile = file.path(dataPath, dataLoadScriptName),
                            categoricalQuestions = c('informedConsent',
                                                     'gender',
                                                     'hasJob',
                                                     'currentEducation',
                                                     'prevEducation',
                                                     'country'));

### Remove one case threatening data integrity
dat <- dat[is.na(dat$age) | dat$age < 100, ];

### Set age to NA for those who did not want to provide their age
dat$age <- ifelse(dat$age==0, NA, dat$age);

### Rename one inconsistently named variable
names(dat)[names(dat) == 'highDose_AttBeliefs_intensity'] <- 'highDose_AttBeliefs_intens';

### Create a set of three regular expressions to add
### underscores behind the behaviors
varnameRegExPairs <- lapply('highDose', function(curBehav) {
  return(c(paste0("^", curBehav, "(.+)"),
           paste0(curBehav, "_\\1")));
});

### Process the variables labels (subquestions, anchors etc)
dat.labelDf <- processLSvarLabels(dat, varnameRegExPairs = varnameRegExPairs);

### Replace variable names with the versions with the
### inserted underscores
names(dat) <- dat.labelDf$varNames.cln;

### Specify the determinant structure
detStruct <-
   determinantStructure('highDose',
                        list('highDose',
                             behaviorRegEx = 'highDose'),
                             determinantVar("intention",
                                            "IntentionRAA",
                                            determinantVar("attitude",
                                                           "AttGeneral",
                                                           subdeterminants("Likelihood",
                                                                           "AttBeliefs"))));

### Process the determinant structure, eventually ordering the
### determinant importance plots

### Add variable names to the determinant structure object
detStructAddVarNames(detStruct,
                     names = grep("\\.\\.", names(dat), value=TRUE, invert=TRUE));
### Compute scales
dat <- detStructComputeScales(detStruct, dat);
### Add variable labels
detStructAddVarLabels(detStruct, dat.labelDf);

### Extract variable labels and translate them
subQuestions.original <- detStruct$intention$attitude$Likelihood$subQuestions;
leftAnchors.original <- detStruct$intention$attitude$Likelihood$leftAnchors;
rightAnchors.original <- detStruct$intention$attitude$Likelihood$rightAnchors;

translations <- matrix(
  c("If I use a high dose of ecstasy,\nmy trip is...", "shorter", "longer",
    "If I use a high dose of ecstasy,\nmy trip is...", "more mild", "more intense",
    "If I use a high dose of ecstasy,\nI get...", "much less\nintoxicated", "much more\nintoxicated",
    "A high dose of ecstasy gives me...", "much less\nenergy", "much more\nenergy",
    "With a high dose of ecstasy,\nthe happy, euphoric feeling is...", "much weaker", "much stronger",
    "If I use a high dose of ecstasy,\nI learn...", "much less\nabout myself", "much more\nabout myself",
    "If I use a high dose of ecstasy,\nI feel...", "much less\nconnected\n to others", "much more\nconnected\nto others",
    "If I use a high dose of ecstasy,\nI can make contact with others...", "much harder", "much easier",
    "If I use a high dose of ecstasy,\nI feel like having sex...", "much less", "much more",
    "If I use a high dose of ecstasy,\nI forget my problems...", "much slower", "much faster",
    "If I use a high dose of ecstasy,\nI feel...", "much less\nself-absorbed", "much more\nself-absorbed",
    "With a high dose of ecstasy,\nI can test my boundaries...", "much less", "much better",
    "If I use a high dose of ecstasy,\nthe music sounds...", "much worse", "much better",
    "If I use a high dose of ecstasy,\nI hallucinate...", "much less", "much more",
    "If I use a high dose of ecstasy,\nI feel time passes...", "much slower", "much faster",
    "If I use a high dose of ecstasy,\nafterwards I remember...", "much less", "much more",
    "For my health, using a high\ndose of ecstasy is...", "much worse", "much better",
    "My experience with using a high\ndose of ecstasy is...", "worse", "better",
    "Do you worry as much, more, or less\nabout the physical side effects of\nusing a high dose of ecstasy?", "worry\nmuch less", "worry\nmuch more",
    "Do you worry as much, more, or less\nabout the emotional and psychic side\neffects of using a high dose of ecstasy?", "worry\nmuch less", "worry\nmuch more",
    "Do you have as much, more, or less\nregret after using a high dose of ecstasy?", "much less\nregret", "much more\nregret"), ncol=3, byrow=TRUE);

subQuestions.translated <- translations[, 1];
leftAnchors.translated <- translations[, 2];
rightAnchors.translated <- translations[, 3];

subQuestions <- subQuestions.translated;
leftAnchors <- leftAnchors.translated;
rightAnchors <- rightAnchors.translated;

selectedVariables <- 14:18;
selectedVarNames <- unlist(detStruct$intention$attitude$Likelihood$varNames)[selectedVariables];

selectedTargets <- c(detStruct$intention$attitude$scaleVarName,
                     detStruct$intention$scaleVarName);

varLabels <- c("A high dose causes more hallucinations",
               "A high dose makes time pass faster",
               "A high dose improves my memory",
               "A high dose is better for my health",
               "A high dose yields a better experience");

### Set 'other' gender to NA
dat$gender <- factor(ifelse(as.numeric(dat$gender) > 1 &
                            as.numeric(dat$gender) < 4,
                            dat$gender, NA),
                     labels=c('Male', 'Female'));

allRelevantVariables <- c(selectedVarNames, selectedTargets, 'gender');

### Remove all irrelevant variables and all missing cases
dat <- dat[complete.cases(dat[, allRelevantVariables]), allRelevantVariables];

```

The examples are based on `r nrow(dat);` participants.

### Challenges when working with confidence intervals

The beliefs selected for this example are `r vecTxt(selectedVarNames, lastDelimiter = ", and ")`. The determinants with which the associations of these beliefs will be computed are `r vecTxt(selectedTargets, lastDelimiter = " and ")`.

```{r tab.cap="Correlation coefficients and their 95% confidence intervals for the associations between the beliefs and the direct measure of attitude, all regarding using a high dose of ecstasy." }

  ########################################################################
  ### Generate table 1
  ########################################################################

  meansVector <- vecTxt(round(colMeans(dat[, selectedVarNames]), 2), lastDelimiter = ", and ");
  sdsVector <- vecTxt(round(apply(dat[, selectedVarNames], 2, sd), 2));
  assocMatrix <- associationMatrix(dat, selectedVarNames, selectedTargets);
  rMatrix <- assocMatrix$output$raw$es;
  rAttitude <- formatR(rMatrix[, 1]);
  rIntention <- formatR(rMatrix[, 2]);

  ciTable <- cbind(formatR(assocMatrix$output$raw$ci.lo[, 1]),
                   formatR(assocMatrix$output$raw$es[, 1]),
                   formatR(assocMatrix$output$raw$ci.hi[, 1]));

  rownames(ciTable) <- varLabels;
  colnames(ciTable) <- c('lower', 'r', 'upper');

  pander(ciTable);

### For easy creation of 'manually specified diamond plots'
# cat0("c(", vecTxtQ(selectedVarNames), ")");

```

For example, for the five beliefs related to using a high dose of ecstasy, the respective means are `r meansVector;`, and their correlations with attitude are `r rAttitude;`. This list of five values is quite straightforward to evaluate. However, when complemented with the lower and upper bounds of the confidence intervals, comparing these estimates requires evaluating fifteen values simultaneously, which carries a considerably higher load (see Table 1). For more associations, this number quickly increases. It would be useful if a method was available to facilitate interpretation of multiple confidence intervals simultaneously.

```{r echo=TRUE, fig.cap="A diamond plot created from means, standard deviations, and sample sizes." }
meanSDtoDiamondPlot(means = c(5.44, 5.58, 2.85, 2.07, 3.76),
                    sd = c(1.19, 1.42, 1.4, 1.14, 1.63),
                    ns = c(201, 201, 201, 201, 201),
                    labels = c('Hallucination',
                               'Time',
                               'Memory',
                               'Health',
                               'Experience'));
```


  



```{r echo=TRUE, fig.cap="The same diamond plot, but with an adjusted x axis." }
meanSDtoDiamondPlot(means = c(5.44, 5.58, 2.85, 2.07, 3.76),
                    sd = c(1.19, 1.42, 1.4, 1.14, 1.63),
                    ns = c(201, 201, 201, 201, 201),
                    labels = c('Hallucination',
                               'Time',
                               'Memory',
                               'Health',
                               'Experience')) +
  scale_x_continuous(limits=c(1, 7), breaks=1:7);
```


  


```{r echo=TRUE, fig.cap="A simple diamond plot based on raw data, but hiding the raw datapoints." }
meansDiamondPlot(dat, items=c('highDose_AttBeliefs_hallucinate',
                              'highDose_AttBeliefs_timeAwareness',
                              'highDose_AttBeliefs_memory',
                              'highDose_AttBeliefs_health',
                              'highDose_AttBeliefs_better'),
                 showData=FALSE);
```



```{r echo=TRUE, fig.cap="The same diamond plot, but ." }

meansDiamondPlot(dat, items=c('highDose_AttBeliefs_hallucinate',
                              'highDose_AttBeliefs_timeAwareness',
                              'highDose_AttBeliefs_memory',
                              'highDose_AttBeliefs_health',
                              'highDose_AttBeliefs_better'));

meansDiamondPlot(dat, items=c('highDose_AttBeliefs_hallucinate',
                              'highDose_AttBeliefs_timeAwareness',
                              'highDose_AttBeliefs_memory',
                              'highDose_AttBeliefs_health',
                              'highDose_AttBeliefs_better'),
                 showData=FALSE) +
  xlim(1, 7);


meansDiamondPlot(dat, selectedVarNames, labels=varLabels);

biAxisDiamondPlot(dat, selectedVarNames,
                  leftAnchors[selectedVariables],
                  rightAnchors[selectedVariables],
                  subQuestions=subQuestions[selectedVariables]);




associationsDiamondPlot(dat, selectedVarNames, selectedTargets[1]);

associationsDiamondPlot(dat, selectedVarNames,
                        selectedTargets[1],
                        alpha=.5, color = 'blue') +
  associationsDiamondPlot(dat, selectedVarNames,
                          selectedTargets[2],
                          alpha=.5, color='red',
                          returnLayerOnly = TRUE);

associationsDiamondPlot(dat, selectedVarNames, 'gender');
associationMatrix(dat, selectedVarNames, 'gender');

```
